---
title: "PS270 Final Project"
author: "Azalea Steffes"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Hypothesis

Is regime change from democracy to authoritarian government accompanied by a in primary school enrollment? By looking at countries who have transitioned from democratic to authoritarian governments I want to analyze how this transition may correlate with decreasing investment in education by the government. By looking at data on education and regime change from three countries: Nicaragua, Venezuela,and Hungary I want to analyze how lowered investment in education is a key indicator of regime change and democratic backsliding. Nicaragua, Venezuala and Hungary are all unique because they have all recently changed from Democratic to authoritarian regimes. Freedom to get an education is an important indicator of how democratic a country is but in my research I want to specifically analyze how states lack of investment in education contributes to democratic backsliding and regime change. The first data set I’m using contains information about every country and investment in education and the second data set I’m using contains information about regime changes democratic back sliding and democracy. Investment in education is measured as a percentage of each country's GDP. The second data set measures democracy on a dichotomous scale. Countries who cross a certain threshold are measured as autocracies and this threshold is determined by a country holding open elections. I will track countries' investment in education in years leading up to a regime change and after a regime change to see how government investment in education correlates with regime change. For my hypothesis to be supported by the data, education spending should decrease significantly in years leading up to regime change and continue decreasing after the regime change occurs. My hypothesis would be disproved if a country's funding for education increased before a regime change occurred and continued to increase afterwards.

## Loading in the Data

```{r, include=FALSE}
library(readr)
library(tidyverse)
library(tidyr)
education <- read_csv("~/Data/world-education-data 2.csv")
glimpse(education)
Regime_chng <-  read_csv("~/Data/rgj2010.csv") 
glimpse(Regime_chng)


```

```{r , include=FALSE}
Data <-
  inner_join(education, Regime_chng,
         by = c("year", "country_code" = "pitfcode"))
```


```{r include=FALSE}
countries <- c("NIC", "VEN", "HUN" )
VEN <- Data |>
 filter(country_code %in% "VEN")
```

```{r}
library(tidyverse)
VEN_diagram <- VEN |>
  ggplot(mapping = aes(x=  year ,
                y = school_enrol_primary_pct ,
                fill= country_code, color = country_code)) +
  labs(x= "year",
       y = "primary School Enrollment ",
       title = "Primary School Enrollment after Regime Change in Venezuala") +
  geom_point() +
  geom_smooth() +
  geom_vline(xintercept = 2005, color = "blue3")+
  theme_minimal(base_size = 13)+
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 8))
VEN_diagram
```


```{r, include = F}
SEN <- Data |>
 filter(country_code %in% "SEN")
```

```{r}
diagram_1 <- SEN |>
  ggplot(mapping = aes(x=  year ,
                y = school_enrol_primary_pct,
                fill= country_code, color = country_code)) +
  labs(x= "year",
       y = "primary School Enrollment",
       title = "Primary School enrollmentafter Regime Change Sengal") +
  geom_point() +
  geom_smooth() +
  geom_vline(xintercept = 2007, color = "green3") +
  theme_minimal(base_size = 13)+
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 8))
  
diagram_1
```
## Sengal and Venezuala
These three graphs visualize how regime change impacts primary school enrollment by looking at these two countries there are no assumptions we can make about how authoritarian regime changes effect primary school enrollment because the sample size is to small however by looking at patterns in more countries we can start to make causal assumptions.


## Prepping The Data

```{r, include=FALSE}
trans <- Data |>
  filter(country_code %in% c("SEN", "COM", "ECU", "FJI", "GNB", "IRQ", "KEN", "MAC", "PAK", "PNG", "RUS", "SEN", "VEN"))
trans
```


```{r, include=FALSE}
Data <- Data |>
  group_by(country_code) |>
  mutate(change_year = year[rgjtrans == -1] [1],
         event_time = year - change_year) |>
  ungroup()

align <- Data |>
        filter (event_time >= -5 & event_time <= 5)
align |>
  select(country_code, year, rgjtrans, change_year, event_time)
```




## Visualizing primary school enrollment before and after an Authoritarian regime change

Trying something

```{r}
full_plot <- Data |>
  filter(!is.na(event_time)) |>
  filter(event_time >= -5 & event_time <= 5) |>
  ggplot(mapping = aes(x = event_time,
                       y = school_enrol_primary_pct,
                       color = country_code)) +
  geom_line(linewidth = 0.7)+
  geom_point(size = 1.5, alpha = 0.4, na.rm = TRUE) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Primary School Enrollment And Regime Change",
       x = "Years To Regime Change",
       y = "Primary School Enrollment") +
    theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 8))
full_plot
```

## Interpretation

These graphs show what individuals countries primary school enrollment rate look like before and after regime change however there is not very clear patterns within while some countries primary school enrollment goes up drastically before an authoritarian regime change others enrollment shoots down. In the After graph again there is not a super clear visual pattern countries have different patterns however there is more data on countries after the regime change and more countries to analyze. Overall in the after graph primary school enrollment rates seem sightly more stable.


## Visualising pattern before and after regime change

```{r}
ba_enroll_chrt <- Data |> 
  drop_na(event_time)|>
  filter(event_time >= -5 & event_time <= 5) |>
  mutate(
    before_after = if_else(event_time < 0, "Before","After"))|>
  group_by(before_after)|>
  summarise(mean_enroll = mean(school_enrol_primary_pct, na.rm= TRUE))|>
  pivot_wider(
    names_from= "before_after",
    values_from = "mean_enroll"
  ) |> 
  mutate(
  ATE = After - Before)
ba_enroll_chrt|>
select(Before, After, ATE)|>
knitr::kable(
  caption= "Primary school enrollment Before and after regime change",
digits=2, 
col.names = c("Before", "After", "ATE")) 
ba_enroll_chrt
```
 

### Linear Regression

```{r}
Data <- Data |>
  filter(!is.na(event_time)) |>
  filter(event_time >= -5 & event_time <= 5) |> 
  mutate(post = if_else(event_time >= 0, 1, 0))

fit <- lm(school_enrol_primary_pct ~ post, data = Data)

summary(fit)

fit |>
  broom::tidy()|>
  select(term, estimate) |>
  knitr:: kable(digits = 2)

```



## Interpretation
*Before* authoritarian regime changes countries have higher rates of primary school enrollment was 2.66 percentage points higher than *after* a country has been taken over by an authoritarian government. When running a regression holding constant for event time frame we see that when event time is held at 0 for before regime change the average primary school enrollment is 97.39. When we look at the post regime change 1 primary school enrollment decreases by 5.02.

## Code Printout

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}
```
